Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-26 mr-review.Rmd
rees_robust_2019 text added

@windmeijer_use_2019, building on the research from 
@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
after recognizing that sisVIVE misclassifies valid SNP instruments as 
invalid when the invalid SNP instruments have strong effects on the 
exposure. Consistent selection of invalid SNP instruments, they found, 
depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
proposed a median estimator with consistency that doesn't 
depend on SNP instrument-exposure association strength or the SNP instrument 
correlation structure. They then applied methods from @zou2006adaptive to 
achieve a consistent estimator with the same distribution as the oracle 
two-stages least squares.

@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for MR analysis with GWAS summary data. 
They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
distribution is the simple mode-based estimate of the causal effect 
[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
weighting the empirical distribution of ratio estimates by the inverse variance of each 
estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators. @hartwig_robust_2017

@@

@oconnor_distinguishing_2018

>>>>

@windmeijer_use_2019, building on the research from 
@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
after recognizing that sisVIVE misclassifies valid SNP instruments as 
invalid when the invalid SNP instruments have strong effects on the 
exposure. Consistent selection of invalid SNP instruments, they found, 
depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
proposed a median estimator with consistency that doesn't 
depend on SNP instrument-exposure association strength or the SNP instrument 
correlation structure. They then applied methods from @zou2006adaptive to 
achieve a consistent estimator with the same asymptotic distribution as the oracle 
two-stages least squares.

@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. 
They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
distribution is the simple mode-based estimate of the causal effect 
[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
weighting the empirical distribution of ratio estimates by the inverse variance of each 
estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.

@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
developing three new methods for two-sample MR analysis with GWAS summary data: robust 
regression, penalized weights, and LASSO penalization. The first two can be viewed as 
modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
downweighting or excluding variants with hetero-
geneous causal estimates.
Recall that MR-Egger offers consistent causal effect estimates when there is no 
horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
formulate an estimator with high breakdown point and high efficiency. The second estimation 
method, with penalized weights, is much like the penalized median estimator, from above, 
except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.

\begin{equation}
\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
(\#eq:lasso)
\end{equation}

@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
and a cross-validation rule. Taken together, these three methods offer additional contributions to 
a suite of sensitivity analysis tools that practitioners should consider.

@@
